\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2023-09-15 docs/papers/sto/so.tex
FIXUP draft

The differentiable cosmological simulation with adjoint method \citep{Li2022a}
and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
foundation of the neural optimization discussed in this work.
Within this fully differentiable framework, we are able to embed neural networks
in any physical process and train them for many purposes, including the
optimization of the gravitational force evaluation in this work.

Following these two works, we denote the particles' state with $\vz = (\vx,
\vp)^\intercal$ and their adjoint variables with $\vlambda = (\vxi, \vpi)$.
We refer to the quantities that are directly related to the final objective
function $\cJ$ as the observables.
In this work, our observable $\vo$ includes multiple snapshots at given times.
For the observation operator $O_{i}^{i-1}$ in the adjoint equations from time
step $a_i$ back to $a_{i-1}$, the adjoint variables are updated through
\begin{align}
  \vxi_{i-1} &:= \vxi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
  \vpi_{i-1} &:= \vpi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
\end{align}
where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
be non-zero if they were used in the interpolation of any snapshots in $\vo$.
The initial conditions are
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
\end{align}
which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$.
The derivative of the objective function with respect to the observable and the
VJP of the observation funciton (mainly the interpolation function discussed
below) can be conveniently computed by AD.

@@

The N-body time integration is carried out following discrete time steps.
Given the particle displacements and velocities at two consecutive time steps
$a_i$ and $a_{i+1}$, we get the status of particles in between at $a$ using the
cubic Hermite interpolation.
The interpolated particle displacements are given by
\begin{align}
  \vx(a) =\ &h_{00}(\alpha)\vx_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
           &h_{01}(\alpha)\vx_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
\end{align}
where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis functions read
\begin{align}
  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1, \nonumber\\
  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha, \nonumber\\
  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2, \nonumber\\
  h_{11}(\alpha) &= \alpha^3 - \alpha^2.
\end{align}
The particle velocities are then given by the derivative $\d\vx / \d a$.
Note that the interpolation is a linear combination of the particles at two
consecutive time steps, which therefore can be done in either two steps or just
one step by carrying the particles from the previous time step.
We find the two-steps scheme more convenient for the backward evolution of the
adjoint variables.

>>>>

The differentiable cosmological simulation with adjoint method \citep{Li2022a}
and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
foundation of the gradient-based optimization discussed in this work.
Within this fully differentiable framework, we are able to embed neural networks
in physical processes and train them for many purposes, including the
optimization of spatial resolution by sharpening the gravitational force in this
work.

Here we briefly recap the equations that are closely related to our
extensions to the original framework, and setup the notations of the variables.
Following \cite{Li2022a}, we denote the particles' state with $\vz = (\vx,
\vp)^\intercal$ and their adjoint variables with $\vlambda = (\vxi, \vpi)$.
We refer to the quantities that are directly related to the final objective
function $\cJ$ as the observable $\vo$.
In this work, our observable includes multiple snapshots at given times.
For the observation operator $O_{i}^{i-1}$ in the adjoint equations (check
Eq.(18) in \cite{Li2022a}) from time step $a_i$ back to $a_{i-1}$, the adjoint
variables are updated through
\begin{align}
  \vxi_{i-1} &:= \vxi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
  \vpi_{i-1} &:= \vpi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
\end{align}
where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
be non-zero only if they were used in the generation of any snapshots in $\vo$.
The initial conditions of these adjoint variables are given by
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
\end{align}
which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$.
The derivative of the objective function with respect to the observable
($\p\cJ/\p\vo$) and the VJPs of the observation operation ($\p\vo/\p\vx_n$ and
$\p\vo/\p\vp_n$, mainly the interpolation function discussed in
\autoref{sec:snapobs} below) can be conveniently computed by AD.

@@

The N-body time integration is carried out following discrete time steps. Given
the particle displacements $\vx$ and velocities $\vv = \vp/(a^3H)$ at two
consecutive time steps $a_i$ and $a_{i+1}$, we get the snapshot of particles in
between at $a$ using
the cubic Hermite interpolation.
The interpolated particle displacements are given by
\begin{align}
  \vx(a) =\ &h_{00}(\alpha)\vx_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
           &h_{01}(\alpha)\vx_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
\end{align}
where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis functions read
\begin{align}
  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1, \nonumber\\
  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha, \nonumber\\
  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2, \nonumber\\
  h_{11}(\alpha) &= \alpha^3 - \alpha^2.
\end{align}
The particle velocities are then given by the derivative $\d\vx / \d a$.
Note that the interpolation is a linear combination of the particle snapshots at
two consecutive time steps, which therefore can be done in either two steps or
just one step by carrying the particles from the previous time step.
We find the two-steps scheme more convenient for the backward evolution of the
adjoint variables.
The interpolated snapshots at different times make the observable in this work,
and the VJPs of the interpolation operation are computed by AD and put into the
adjoint equations as mentioned in \autoref{sec:pmwd}.
